A Transactional Memory System for Parallel Python

### 2014-06-06 talk/dls2014/paper/paper.tex
revised conclusions

\begin{abstract}
....
\end{abstract}

@@

\keywords
...

@@

\subsection{Application View}
\remi{possibly adapt to application view \& benefits (atomic blocks)}
\remi{XXX: Also, make own section or integrate with introduction}

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter in order to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We follow the direct approach of using TM instead of the
GIL.

@@

Our TM system is fully implemented in software. However, we do exploit
some more advanced features of current CPUs, particularly \emph{memory
segmentation, virtual memory,} and the 64-bit address space. Still,
it cannot be classified as a hybrid TM system since it currently
makes no use of any HTM present in the CPU.

@@

Our STM system fundamentally builds on threads. Transactions run in
their context, and there can be only one transaction per thread at a
time. Transactions run in parallel because these threads can run in
parallel. In order to isolate transactions from each other, we look
at isolating the memory accesses from threads.

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack}~\cite{fergus02} in order to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!.

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to.

@@

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \textbf{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).

@@

A requirement for performing a commit is to synchronise all threads so
that we can safely update objects in other segments. To make this
synchronisation fast and cheap, we do not want to insert an additional
check regularly in order to see if synchronisation is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \lstinline!stm_allocate!  shown
below:

@@

This code does simple bump-pointer allocation in the nursery. If there
is still space left in the nursery, we return
\lstinline!nursery_current!  and bump it up by
\lstinline!size_rounded!.  The interesting part is the check
\lstinline!nursery_current > nursery_end!  which will trigger the slow
path of the function to possibly perform a minor collection in order
to free up space in the nursery.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
  in order to let long transactions succeed more often
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to~\cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

We use atomic blocks in order to easily synchronise access to shared
data structures in the first three benchmarks. For the embarrassingly
parallel benchmarks, we do not need any explicit synchronisation at
all. These atomic blocks are simulated with a single global lock
when running on top of GIL-supported interpreters. With our STM
system, they map to transactions that will execute optimistically
in parallel.

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones (\remi{$avg=2.5\times$} speedup) and a little less for
the others (\remi{$avg=1.9\times$} speedup). The reason for that is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks in order do synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

@@

The speedups from enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, in order to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times (denoted with ``(large)'' in benchmark names).

@@

Overall PyPy needs the JIT in order for its performance to be
competitive with CPython's. It would be interesting to see how using
our STM system in CPython would turn out, but it is a lot of work. On
its own, our system scales well, so we hope to also see that with the
JIT in the future.

@@

% review of paper
We set out to resolve the issue of parallelism in Python by replacing
the GIL with TM. In particular we deemed a close integration of STM
and the GC best as a way to attack the issues of atomicity,
synchronisation, and memory management in a unified way. While the
interplay of STM and the GC gives us some important performance
benefits, we also implemented atomic blocks backed by STM for Python
as a scalable synchronisation mechanism.

% STM needs good performance:
Since in our approach, STM becomes a vital part of the interpreter,
good performance is essential. We found an interesting way to combine
the existing memory segmentation and virtual memory features of
current CPUs to lower the overhead of STM. We want to stress the
importance of memory segmentation to our approach, because it seems
rather unvalued. In the move from the x86 to the x86{-}64
architecture, they already removed 4 out of 6 segment registers.  We
want to make the point that there are still valid and interesting uses
for them (see also \cite{bennet10}).

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement yields an average speedup of \remi{$2.5\times$} for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reached \remi{$1.9\times$}.

For the overall performance compared to our baseline interpreter
CPython, integration with our JIT compiler is necessary. Once this
integration matures, we believe our approach to not only serve as a
simple GIL replacement, but also as a way forward in parallelising
languages -- making atomicity and synchronisation the language's
instead of the programmer's problem.

@@

\acks
...

@@

\bibitem{beazley10} Beazley, David. "Understanding the python gil."
  \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.

@@

\bibitem{odaira14}
  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
  global interpreter locks in ruby through hardware transactional
  memory. In \emph{Proceedings of the 19th ACM SIGPLAN symposium on
    Principles and practice of parallel programming} (PPoPP '14).

@@

\bibitem{fuad10}
  Fuad Tabba. 2010. Adding concurrency in python using a commercial
  processor's hardware transactional memory support. \emph{SIGARCH
  Comput. Archit. News 38}, 5 (April 2010), 12-19.

@@

\bibitem{victor11}
  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
  transactional memory vs. locks in practice. In \emph{Proceedings of
    the twenty-third annual ACM symposium on Parallelism in algorithms
    and architectures} (SPAA '11).

\bibitem{christopher10}
  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
  Witchel. 2010. Is transactional programming actually
  easier?. \emph{SIGPLAN} Not. 45, 5 (January 2010), 47-56.

>>>>

\begin{abstract}
\remi{missing abstract}
\end{abstract}

@@

\keywords
\remi{missing keywords}

@@

\subsection{Application View}

@@

We focus on the \emph{threading} approach. This requires us to remove
the GIL from the interpreter to run code in parallel on
multiple threads. One approach to this is fine-grained locking instead
of a single global lock. Jython~\cite{webjython} and
IronPython~\cite{ironpython} follow this approach. Fine-grained
locking is, however, not a \emph{direct} replacement for the GIL. It
requires multiple locks in multiple places, not only in a central
location. We follow the direct approach of using TM instead of the
GIL.

@@

Our TM system is fully implemented in software. However, we exploit
some more advanced features of current CPUs, particularly \emph{memory
segmentation, virtual memory,} and the 64-bit address space. Still,
it cannot be classified as a hybrid TM system since it currently
makes no use of any HTM present in the CPU.

@@

Our STM system fundamentally builds on threads. Transactions run in
their context, and there can be only one transaction per thread at a
time. Transactions run in parallel because these threads can run in
parallel. To isolate transactions from each other, we look
at isolating the memory accesses from threads.

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack}~\cite{fergus02} to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!.

@@

The \textbf{slow path} is shown in Listing~\ref{lst:w_slow}.
First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection to mark the young object
alive and to update its reference to the new location it gets moved
to.

@@

For the \emph{STM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
is a simple, \textbf{global} array of bytes that is indexed with the
$SO$ of the object divided by 16. Note, ``global'' here really means
it is a single array with data for all segments, there is no address
translation going on to access its elements contrary to e.g.\ the
read markers array.  If we already own the lock, we are done.
If someone else owns the lock, we perform  write-write contention
management that aborts either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we perform contention management to either abort us or
the other transaction, or to simply wait a bit (see Section~\ref{subsub:contentionmanagement}).

@@

A requirement for performing a commit is to synchronise all threads so
that we can safely update objects in other segments. To make this
synchronisation fast and cheap, we do not want to insert an additional
check regularly to see if synchronisation is requested. We
use a trick relying on the fact that dynamic languages are usually
very high-level and thus allocate a lot of objects very regularly.
This is done through the function \lstinline!stm_allocate!  shown
below:

@@

This code does simple bump-pointer allocation in the nursery. If there
is still space left in the nursery, we return
\lstinline!nursery_current!  and bump it up by
\lstinline!size_rounded!.  The interesting part is the check
\lstinline!nursery_current > nursery_end!  which will trigger the slow
path of the function to possibly perform a minor collection
to free up space in the nursery.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
  to let long transactions succeed more often
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to~\cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

We use atomic blocks to easily synchronise access to shared
data structures in the first three benchmarks. For the embarrassingly
parallel benchmarks, we do not need any explicit synchronisation at
all. These atomic blocks are simulated with a single global lock
when running on top of GIL-supported interpreters. With our STM
system, they map to transactions that will execute optimistically
in parallel.

@@

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones (\remi{$avg=2.5\times$} speedup) and a little less for
the others (\remi{$avg=1.9\times$} speedup). The reason for that is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

@@

The speedups from enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times (denoted with ``(large)'' in benchmark names).

@@

Overall PyPy needs the JIT  for its performance to be
competitive with CPython's. It would be interesting to see how using
our STM system in CPython would turn out, but an in-depth evaluation is
beyond the scope of this paper. On
its own, our system scales well, so we expect to also see this trend in the presence of a
JIT in the future.

@@

% review of paper
As multi-core processors continue to increase the number of cores per
processor, it is important that dynamic languages like Python embrace
parallelism.  We report here an effort to replace the Global
Interpreter Lock (GIL) used in many Python implementations with
transactional memory (TM). The key factor for success is a close
integration of a software TM (STM) and the garbage collector to
provide atomicity, synchronisation, and memory management in a unified
way.  The STM system supports atomic blocks as a user abstraction and
thereby provides a scalable synchronisation mechanism to Python
applications.

% STM needs good performance:
The solution presented here is based on a software TM system.  As the
STM is a vital part of the interpreter, good performance is
essential. We present an interesting way to combine the existing
memory segmentation and virtual memory features of current CPUs to
lower the overhead of STM. This software implementation reduces
dramatically the cost of transactional memory and at the same time
provides an opportunity to integrate the STM system with garbage
collection.  This implementation leverages the processor's memory
segmentation hardware, which has lately played only a minor role in
our operating system.  Unfortunately, processor architects seem to
undervalue this hardware resource -- in the evolution from the x86 to
the x86{-}64 architecture, 4 out of 6 segment registers have been
removed. This development potentially threatens the STM strategy
outlined here and we want to remind the architecture community that
this unit can support interesting implementations -- a viewpoint that
has also been expressed by other researchers (see also
\cite{bennet10}).

% results, outlook
The early results presented here are very encouraging. STM as a simple
GIL replacement yields an average speedup of \remi{$2.5\times$} for
embarrassingly parallel workloads on 4 threads.  When also used for
synchronisation in the form of atomic blocks, the average speedup
still reaches \remi{$1.9\times$}.

To obtain an overall performance that is competitive with the
best-performing Python systems (Jython, PyPy), integration of the
STM-based approach with a JIT compiler is necessary. Once this
integration matures, the approach outlined here serves not only as a
simple GIL replacement but also provides a way forward towards
parallelizing Python.

@@

\acks
\remi{cfbolz, felber, marlier, fäs, mastoras}

@@

\bibitem{beazley10} Beazley, David. "Understanding the Python GIL."
  \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.

@@

\bibitem{odaira14}
  Rei Odaira, Jose G. Castanos, and Hisanobu Tomari. 2014. Eliminating
  global interpreter locks in ruby through hardware transactional
  memory. In \emph{Proc. 19th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '14).

@@

\bibitem{fuad10}
  Fuad Tabba. 2010. Adding concurrency in Python using a commercial
  processor's hardware transactional memory support. \emph{SIGARCH
  Comput. Archit. News 38}, 5 (April 2010), 12-19.

@@

\bibitem{victor11}
  Victor Pankratius and Ali-Reza Adl-Tabatabai. 2011. A study of
  transactional memory vs. locks in practice. In \emph{Proc.
     twenty-third annual ACM Symposium on Parallelism in Algorithms
    and Architectures} (SPAA '11).

\bibitem{christopher10}
  Christopher J. Rossbach, Owen S. Hofmann, and Emmett
  Witchel. 2010. Is transactional programming actually
  easier?. \emph{Proc. 15th ACM SIGPLAN Symposium on
    Principles and Practice of Parallel Programming} (PPoPP '10).
